---
title: "courseworkQ2a"
author: "Chris_Pang"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## For Coursework Q, Student ID: 220457882

Creating databse. Cleaning and preparing datas
```{r database}

library(DBI)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggthemes)
# set the working directory that contains the files
setwd("/Users/dataverse_files")

# ======== create the database ========
if (file.exists("airline2.db")) 
  file.remove("airline2.db")
conn <- dbConnect(RSQLite::SQLite(), "airline2.db")



# ======== write to the database ========
# load in the data from the csv files
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)
dbWriteTable(conn, "airports", airports)
dbWriteTable(conn, "carriers", carriers)
dbWriteTable(conn, "planes", planes)

for(i in c(1991:2000)) {
  ontime <- read.csv(paste0(i, ".csv"), header = TRUE)
  if(i == 2000) {
    dbWriteTable(conn, "ontime", ontime)
  } else {
    dbWriteTable(conn, "ontime", ontime, append = TRUE)
  }
}

```


Answering Q2a, plotting the graph individuallly from 1991 to 1999.
```{r Q2a}
#==================Q2a Best time of the day to fly throught 1991 to 1999============================
library(DBI)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
time_segment <- function(deptime) {
  hour <- as.numeric(substr(deptime, 1, 2))
  
  if (hour >= 2 && hour < 6) {
    return('Early Morning')
  } else if (hour >= 6 && hour < 10) {
    return('Morning')
  } else if (hour >= 10 && hour < 14) {
    return('Midday')
  } else if (hour >= 14 && hour < 17) {
    return('Afternoon')
  } else if (hour >= 17 && hour < 21) {
    return('Evening')
  } else {
    return('Night')
  }
}

# Initialize a list to store each year's lowest delay count and corresponding segments
lowest_delay_counts_by_year <- list()

for (year in 1991:1999) {
  query <- sprintf("SELECT CAST(DepTime AS INTEGER) AS DepTime FROM ontime WHERE ArrDelay > 15 AND DepDelay > 15 AND Year = %d", year)
  timeOfDayDf <- dbGetQuery(conn, query)
  timeOfDayDf$DepTime <- sprintf("%04d", timeOfDayDf$DepTime)
  timeOfDayDf$Segment <- sapply(timeOfDayDf$DepTime, time_segment)
  
  segment_order <- factor(c('Early Morning', 'Morning', 'Midday', 'Afternoon', 'Evening', 'Night'), 
                          levels = c('Early Morning', 'Morning', 'Midday', 'Afternoon', 'Evening', 'Night'))
  timeOfDayDf$Segment <- factor(timeOfDayDf$Segment, levels = levels(segment_order))
  
  segment_counts_ordered <- as.data.frame(table(timeOfDayDf$Segment))
  colnames(segment_counts_ordered) <- c("Segment", "Count")
  
  # Find the lowest delay count for the year and store it
  min_delay_count <- min(segment_counts_ordered$Count)
  lowest_delay_segments <- segment_counts_ordered$Segment[which.min(segment_counts_ordered$Count)]
  lowest_delay_counts_by_year[[as.character(year)]] <- list(count = min_delay_count, segment = lowest_delay_segments)
  
  # Plotting
  ggplot(segment_counts_ordered, aes(x = Segment, y = Count, fill = Segment)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    geom_text(aes(label = Count), vjust = -0.3, size = 3.5) +
    scale_fill_brewer(palette = "Pastel1") +
    labs(title = sprintf("Count of Flights Delayed by Time of Day in %d", year),
         x = "Time of Day Segment",
         y = "Number of Flights") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    guides(fill = guide_legend(title = "Time Segments", override.aes = list(fill = brewer.pal(6, "Pastel1")))) +
    scale_x_discrete(labels = c("Early Morning: 0200-0600", "Morning: 0600-1000", "Midday: 1000-1400", 
                                "Afternoon: 1400-1700", "Evening: 1700-2100", "Night: 2100-0200"))
  
  print(ggplot2::last_plot()) # Print the plot for each year
} 
  
  
  
#==================Q2a plotting the 9 years collated graph ============================

years <- names(lowest_delay_counts_by_year)
lowest_counts <- sapply(lowest_delay_counts_by_year, function(info) info$count)
segments <- sapply(lowest_delay_counts_by_year, function(info) info$segment)

# Convert counts to numeric
lowest_counts <- as.numeric(lowest_counts)

# Determine the range of y-values to dynamically position text within the graph
y_min <- min(lowest_counts)
y_max <- max(lowest_counts)
y_range <- y_max - y_min
offset <- y_range * 0.02  # Offset for text positioning, 1% of the y-range
plot(years, lowest_counts, type = "o", col = "#008080", xlab = "Year", ylab = "Number of Delayed Flights", main = "Lowest Number of Delayed Flights by Year with Time of Day")
grid()
lines(years, lowest_counts, type = "o", col = "#008080")

#adjusting text size 
text_size <- 0.501 


# Adjust the offset to position text below the points
offset <- -abs(offset)  # Ensure offset is negative to move text below

for (i in 1:length(years)) {
  year <- years[i]
  count <- lowest_counts[i]
  segment <- segments[i]
  
  
  
  vertical_position <- count + offset  # Adjust vertical position to be below the point
  
  # Place text below each point. Adjust 'adj' for fine-tuning horizontal alignment if necessary.
  text(year, vertical_position, labels = paste(segment, "\n(", count, ")", sep=""), col = "darkblue", cex = text_size, adj = c(0.5, 1))
}


```


Answering Q2a,plotting best day of the week to fly individually from 1991-1999
```{r Q2a Day of the week}

#==================Q2a best day of the week to fly to minimize delay ============================
lowest_delay_day_by_year <- list()

for (year in 1991:2000) {
  query <- sprintf("
    SELECT DayOfWeek, COUNT(*) AS Count 
    FROM ontime 
    WHERE Year = %d AND ArrDelay > 15 AND DepDelay > 15 
    GROUP BY DayOfWeek
    ORDER BY DayOfWeek", year)
  
  # Execute the query
  dayOfWeekCounts <- dbGetQuery(conn, query)
  
  # Initialize a dataframe for all days with counts set to 0
  counts <- data.frame(DayOfWeek = 1:7, Count = rep(0, 7))
  
  # Check if any data is retrieved
  if (nrow(dayOfWeekCounts) == 0) {
    print(paste("No data retrieved for year", year))
    next  # Skip this year if no data is returned
  }

  # Update counts based on the query result
  for (row in 1:nrow(dayOfWeekCounts)) {
    # Ensure that 'DayOfWeek' column exists and has valid data
    if ("DayOfWeek" %in% names(dayOfWeekCounts) && !is.na(dayOfWeekCounts[row, "DayOfWeek"])) {
      day <- dayOfWeekCounts[row, "DayOfWeek"]
      if (day >= 1 && day <= 7) {
        counts[day, "Count"] <- dayOfWeekCounts[row, "Count"]
      } else {
        print(paste("Invalid day:", day, "in year", year))
      }
    } else {
      print(paste("Missing or invalid 'DayOfWeek' data in year", year))
    }
  }
  
  # Plotting a bar chart for each year
  tick_label <- c('Sun', 'Mon', 'Tues', 'Wed', 'Thurs', 'Fri', 'Sat')
  ggplot(counts, aes(x = factor(DayOfWeek, labels = tick_label), y = Count)) +
    geom_bar(stat = "identity", fill = "lightblue") +
    geom_text(aes(label = Count), vjust = -0.3, size = 3.5) +
    labs(x = 'Day of the Week', y = 'Number of Delayed Flights', 
         title = sprintf('Delayed Flights by Day of the Week in %d', year)) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 16, face = "bold"))
  
  # Force the plot to display
  print(ggplot2::last_plot())
  
  # Store the day with the least delays for the year
  min_delay_count <- min(counts$Count)
  min_delay_day <- tick_label[which.min(counts$Count)]
  lowest_delay_day_by_year[[as.character(year)]] <- list(Day = min_delay_day, Count = min_delay_count)
}



#==================Q2a  plotting the 9 years collated graph ============================

# Create a data frame from the list
data <- data.frame(
  Year = as.numeric(names(lowest_delay_day_by_year)),
  Lowest_Count = sapply(lowest_delay_day_by_year, function(info) info$Count),
  Lowest_Days = sapply(lowest_delay_day_by_year, function(info) info$Day)
)

# Plotting
ggplot(data, aes(x = Year, y = Lowest_Count)) +
  geom_line(color = "#008080", size = 1) +  # Using hex code for teal
  geom_point(color = "#008080", size = 3) +  # Using hex code for teal
  geom_text(aes(label = paste(Lowest_Days, "\n(", Lowest_Count, ")", sep="")), 
            nudge_y = (max(data$Lowest_Count) - min(data$Lowest_Count)) * 0.03, 
            size = 3, color = "darkblue", 
            hjust = 0.1, vjust = 2,
            check_overlap = TRUE, 
            family = "serif") +
  labs(title = 'Lowest Number of Delayed Flights by Year',
       x = 'Year', y = 'Count of Lowest Delayed Flights') +
  theme_minimal(base_size = 14) + 
  theme(plot.title = element_text(size = 16, face = "bold")) +
  scale_x_continuous(breaks = data$Year) +  # Ensure X axis has all years as breaks
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +  # Add some padding around Y axis
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") 

# Adjust plot theme for better readability
theme_set(theme_minimal(base_size = ))
```
For Q2b, Plotting do older planes suffer more delay graph individually from 1991-1999
```{r Q2b do older planes suffer more delay}

#==================Q2b plotting do older planes suffer more delay on a year to year basis============================

# Initialize an empty list to store the data
max_delay_age_by_year <- list()

for (year in 1991:1999) {
  # Construct the SQL query
  query <- sprintf("
    SELECT CAST(planes.year AS INTEGER) AS year
    FROM planes
    JOIN ontime ON planes.TailNum = ontime.TailNum
    WHERE CAST(ontime.ArrDelay AS INTEGER) > 15 AND CAST(ontime.DepDelay AS INTEGER) > 15
    AND planes.year IS NOT NULL
    AND CAST(ontime.Year AS INTEGER) >= %s
  ", year)
  
  # Execute the query
  dateOfPlane <- dbGetQuery(conn, query)
  
  # Check if data was fetched
  if (nrow(dateOfPlane) == 0) {
    cat(sprintf("No data fetched for %d.\n", year))
    next
  }
  
  # Calculate the age of the plane in that year
  dateOfPlane$`Age in year` <- year - dateOfPlane$year
  
  # Countby age
  age_counts <- count(dateOfPlane, `Age in year`)
  
  # Find the age with the maximum count
  max_age <- age_counts[which.max(age_counts$n),]
  max_delay_age_by_year[[as.character(year)]] <- list(Age = max_age$`Age in year`, Delays = max_age$n)
  
  # Plot the graph for the specific year
  p <- ggplot(age_counts, aes(x = `Age in year`, y = n, fill = `Age in year`)) +
    geom_bar(stat = "identity", color = "skyblue", fill = "skyblue") +
    scale_x_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 5)) +
    labs(title = sprintf("Number of Flights with Delay by Plane Age in %d", year),
         x = "Age of the Plane",
         y = "Number of Flights with More Than 15 Min Delay") +
    theme_minimal() +
    theme(axis.text.x = element_text( vjust = 0.5, hjust=1))
  
  print(p)
}

# Output the results
for (year in names(max_delay_age_by_year)) {
  cat(sprintf("In %s, the age with the most delays was %s years with %s delays.\n",
              year, max_delay_age_by_year[[year]]$Age, max_delay_age_by_year[[year]]$Delays))
}
  


#==================plotting the 9 years collated regression graph============================
library(ggplot2)
library(dplyr)

# Convert the list to a data frame
data <- bind_rows(max_delay_age_by_year, .id = "Year") %>%
  mutate(Year = as.numeric(Year), Age = as.numeric(Age), Delays = as.numeric(Delays)) %>%
  arrange(Year)

# Plotting with annotations
p <- ggplot(data, aes(x = Age, y = Delays)) +
  geom_point(color = "blue", size = 3, show.legend = TRUE, aes(label = Delays)) +
  geom_text(aes(label = Delays), vjust = 1.5, color = "black", size = 3) +
  labs(title = "Plane Age vs Number of Delays with Counts",
       x = "Age of Planes", y = "Number of Delays") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16),
        axis.title = element_text(size = 13)) +
  geom_smooth(method = "lm", color = "red", se = FALSE, show.legend = TRUE) +
  ggtitle("Plane Age vs Number of Delays with Trend Line and Counts") +
  xlab("Age of Planes") + ylab("Number of Delays") +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("blue", "red"), labels = c("Delays vs. Age", "Line of Best Fit"))

print(p)

# Optionally, calculate the correlation coefficient
correlation <- cor(data$Age, data$Delays, use = "complete.obs")
cat(sprintf("Correlation coefficient between plane age and number of delays: %.2f\n", correlation))


```


Plotting for q2c)
```{r Q2c}

library(tidyverse)

set.seed(42)

for (year in 1991:2000) {
  # Subset the data for the current year
  year_data <- filter(ontime, Year == year)
  
  # Impute missing values in features
  imputed_data <- year_data %>%
    select(DepTime, ArrTime, Distance) %>%
    bind_cols(Year = year_data$Year, Diverted = year_data$Diverted) %>%
    replace_na(list(DepTime = mean(year_data$DepTime, na.rm = TRUE),
                    ArrTime = mean(year_data$ArrTime, na.rm = TRUE),
                    Distance = mean(year_data$Distance, na.rm = TRUE))).  
  
  # Splitting the data into features (X) and target (y)
  X <- imputed_data %>% select(DepTime, ArrTime, Distance)
  y <- imputed_data$Diverted
  
  # Splitting dataset into training and testing sets
  trainIndex <- createDataPartition(y, p = .8, list = FALSE, times = 1)
  X_train <- X[trainIndex, ]
  X_test <- X[-trainIndex, ]
  y_train <- y[trainIndex]
  y_test <- y[-trainIndex]

  # Initializing and fitting the logistic regression model
  logistic_model <- train(X_train, y_train, method = 'glm', family = 'binomial', trControl = trainControl(method = 'none'))
  
  # Getting coefficients and intercept
  coefficients <- c(intercept = coef(logistic_model$finalModel)['(Intercept)'], coef(logistic_model$finalModel)[-1])
  
  # Creating a DataFrame to hold coefficients for the current year
  temp_df <- data.frame(t(coefficients), Year = year)
  
  # Adding temporary DataFrame to the main coefficients DataFrame
  coefficients_df <- bind_rows(coefficients_df, temp_df)
}

# Setting 'Year' as the DataFrame index for plotting
coefficients_df$Year <- factor(coefficients_df$Year)

# Plotting the coefficients for each feature across years
coefficients_df %>%
  pivot_longer(cols = -Year, names_to = "Feature", values_to = "Coefficient") %>%
  ggplot(aes(x = Year, y = Coefficient, color = Feature, group = Feature)) +
  geom_line() +
  geom_point() +
  labs(title = 'Logistic Regression Coefficients for Diverted Flights Across Years',
       x = 'Year', y = 'Coefficient Value') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_brewer(palette = "Dark2")


```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
